---
title: "Data Visualization and Tables in R"
subtitle: "R Bootcamp — Day 2"
author: "Joshua Lukemire"
format:
  revealjs:
    theme: simple
    slide-number: true
    slide-level: 1
    incremental: false
    transition: fade
    code-copy: true
execute:
  echo: false
  warning: false
  message: false
---

<style>
.reveal {
  font-size: 26px;
}

.reveal .slides {
  text-align: left;
}

.reveal h1 {
  font-size: 1.5em;
}

.reveal h2 {
  font-size: 1.15em;
}

.reveal h3 {
  font-size: 1.0em;
}

.reveal p,
.reveal li {
  line-height: 1.22;
}

.reveal pre {
  width: 100%;
  margin-top: 0.3em;
  margin-bottom: 0.3em;
}

.reveal pre code {
  font-size: 1.28em;
  line-height: 1.15;
  max-height: 520px;
}

.reveal table {
  font-size: 0.82em;
  width: 100%;
}

.reveal td,
.reveal th {
  padding: 0.25em 0.45em;
}

.callout {
  border-left: 5px solid #666;
  padding-left: 0.7em;
  margin-top: 0.7em;
}

.question {
  border-left: 5px solid #999;
  padding-left: 0.7em;
  margin-top: 0.7em;
  font-weight: 600;
}

.part {
  font-size: 1.15em;
  font-weight: 600;
  letter-spacing: 0.02em;
}
</style>

# Welcome Back

## Data Visualization and Tables in R

Today focuses on turning cleaned and summarized data into figures and tables.

By the end of the full session, you should be able to:

- Build common statistical graphics using `ggplot2`
- Modify existing plot code
- Choose appropriate geoms for common data questions
- Customize labels, themes, facets, and grouped displays
- Create polished tables using `gt`
- Create descriptive summary tables using `gtsummary`

# Today's Roadmap

| Part | Topic |
|---|---|
| Part I | Grammar of Graphics and Your First `ggplot` |
| Part II | Histograms and Bar Charts |
| Part III | Scatterplots, Boxplots, and Density Plots |
| Part IV | Aesthetics |
| Part V | Grouped Plots and Position Adjustments |
| Part VI | Faceting |
| Part VII | Practical Plot Improvements |
| Part VIII | Food Exposure Visualizations |
| Part IX | Common `ggplot` Errors |
| Part X | Tables with `gt` |
| Part XI | Tables with `gtsummary` |
| Part XII | Capstone |

<div class="callout">
We will spend most of the day on `ggplot2`, then close with practical table-making tools.
</div>

# Setup

We will continue using the Oswego dataset.

The examples assume:

- You are working inside an R Project
- Your data are stored in a folder named `Data`
- `here()` points to the folder directly above `Data`

```{r}
#| echo: true
library(tidyverse)
library(here)
library(gt)
library(gtsummary)

oswego <- read_csv(
  here("Data", "oswego.csv")
)
```

# Part I

## Grammar of Graphics and Your First `ggplot`

<div class="part">Understanding how `ggplot2` builds figures</div>

# Part I Learning Goals

By the end of this section, you should be able to:

- Recognize the basic structure of `ggplot` code
- Identify the data source in a plot
- Identify aesthetics inside `aes()`
- Identify the geom being used
- Understand how layers are added with `+`
- Modify a simple plot

# The Core `ggplot` Pattern

Most `ggplot2` figures follow the same basic pattern:

```r
ggplot(data, aes(...)) +
  geom_...
```

or

```r
data %>%
  ggplot(aes(...)) +
  geom_...
```

The three essential pieces are:

| Piece | Purpose |
|---|---|
| Data | The data frame being plotted |
| Aesthetics | Variables mapped to visual features |
| Geom | The type of plot to draw |

# Component 1: Data

The first argument tells `ggplot` where the data live.

```r
ggplot(
  oswego,
  aes(x = age)
)
```

Here:

```r
oswego
```

is the dataset.

Every `ggplot` starts with a data frame.

# Component 2: Aesthetics

Aesthetics determine which variables are connected to visual features.

```r
aes(
  x = age
)
```

This tells `ggplot`:

> Put `age` on the x-axis.

Later we will use aesthetics such as:

```r
x
y
fill
color
group
```

# Component 3: Geometries

Geometries determine what type of plot to draw.

Common geoms include:

| Geom | Plot Type |
|---|---|
| `geom_histogram()` | Histogram |
| `geom_bar()` | Bar chart from raw data |
| `geom_col()` | Bar chart from summarized data |
| `geom_point()` | Scatterplot |
| `geom_boxplot()` | Boxplot |
| `geom_density()` | Density plot |


# The Plus Sign Adds Layers

`ggplot2` builds figures layer by layer.

```r
ggplot(
  oswego,
  aes(x = age)
) +
  geom_histogram()
```

The `+` adds a layer.

Later we will add layers such as:

```r
labs()
theme_classic()
facet_wrap()
```

# Putting It Together

```{r}
#| echo: true
ggplot(oswego, aes(x = age)) +
  geom_histogram()
```

Read this as:

1. Use the `oswego` dataset
2. Put `age` on the x-axis
3. Draw a histogram

# Exercise: Find the Missing Piece

What is missing?

```r
ggplot(oswego) +
  geom_histogram()
```

# Exercise: Modify the Code

Goal:

> Create a histogram of `timesupper` instead of `age`.

Start with:

```r
ggplot(
  oswego,
  aes(x = age)
) +
  geom_histogram()
```

Modify the code.

# Checkpoint: Grammar of Graphics

## You should now be able to:

- Recognize the basic `ggplot()` structure
- Identify the dataset
- Identify variables inside `aes()`
- Identify the geom
- Explain why `+` is used

## Key ideas

- `ggplot2` builds figures from layers.
- Data, aesthetics, and geoms work together.
- A plot usually needs both aesthetics and at least one geom.

<div class="question">
Questions before we move to common plot types?
</div>

# Part II

## Histograms and Bar Charts

<div class="part">Distributions and categorical counts</div>

# Part II Learning Goals

By the end of this section, you should be able to:

- Create histograms for numeric variables
- Change the number of bins in a histogram
- Create bar charts for categorical or logical variables
- Explain what `geom_bar()` computes automatically
- Explain when to use `geom_col()`
- Recognize the difference between raw data and summarized data in plots

# Histograms

Histograms are used for numeric variables.

Common examples:

- Age
- Weight
- Blood pressure
- Time to symptom onset
- Time since exposure

A histogram helps show the distribution of a variable.

# First Histogram

```{r}
#| echo: true
ggplot(
  oswego,
  aes(x = age)
) +
  geom_histogram()
```

This creates a histogram of attendee ages.

Each bar represents a range of ages.

The height of each bar is the number of observations in that range.

# Interpreting Histograms

Histograms can help answer questions like:

- What values are common?
- What values are rare?
- Is the distribution roughly symmetric?
- Are there unusual values?
- Is the distribution concentrated in one region?

For this dataset, an age histogram gives a quick sense of who attended the supper.

# Histogram Bins

Histograms group numeric values into bins.

Default:

```r
geom_histogram()
```

Custom number of bins:

```r
geom_histogram(
  bins = 10
)
```

The number of bins affects the appearance of the plot.

# Fewer Bins

```{r}
#| echo: true
ggplot(oswego, aes(x = age)) +
  geom_histogram(bins = 10)
```

Fewer bins usually produce:

- a smoother appearance
- less detail
- a broader sense of the distribution

# More Bins

```{r}
#| echo: true
ggplot(oswego, aes(x = age)) +
  geom_histogram(bins = 30)
```

More bins usually produce:

- more detail
- more variation across bars
- potentially more noise

# Exercise: Modify the Code

Goal:

> Create an age histogram using 25 bins.

Start with:

```r
ggplot(
  oswego,
  aes(x = age)
) +
  geom_histogram()
```

Modify the code.

# Exercise: Predict the Output

What do you expect to change?

```r
ggplot(
  oswego,
  aes(x = age)
) +
  geom_histogram(
    bins = 50
  )
```

Question:

Will the figure become smoother or more detailed?

# Bar Charts

Bar charts are used for categorical or logical variables.

Common examples:

- Sex
- Illness status
- Exposure status
- County
- Vaccination status

A bar chart shows counts by category.

# First Bar Chart

```{r}
#| echo: true
ggplot(oswego, aes(x = ill)) +
  geom_bar()
```

Because `ill` is logical, the x-axis categories are:

```r
FALSE
TRUE
```

This shows counts of non-cases and cases.

# What Does `geom_bar()` Do?

`geom_bar()` computes counts automatically.

```r
ggplot(
  oswego,
  aes(x = ill)
) +
  geom_bar()
```

is conceptually similar to:

```r
oswego %>%
  count(ill)
```

You do not need to precompute counts before using `geom_bar()`.

# Another Bar Chart

```{r}
#| echo: true
ggplot(oswego, aes(x = sex)) +
  geom_bar()
```

This shows the number of observations in each sex category.

The height of each bar is a count.

# `geom_col()`

`geom_col()` is different.

It expects you to provide both:

- an x variable
- a y variable

Example:

```{r}
#| echo: true
ill_counts <- oswego %>%
  count(ill)

ggplot(ill_counts, aes(x = ill, y = n)) +
  geom_col()
```

# `geom_bar()` vs `geom_col()`

| Function | Data Required | What It Does |
|---|---|---|
| `geom_bar()` | Raw data | Counts rows automatically |
| `geom_col()` | Summarized data | Uses existing y-values |

Use `geom_bar()` when you want `ggplot` to count.

Use `geom_col()` when you already have counts or rates.

# Exercise: Which Geom?

You have this dataset:

```r
jello_counts <- oswego %>%
  count(jello)
```

You want to plot `n` by `jello`.

Should you use:

```r
geom_bar()
```

or

```r
geom_col()
```

Why?

# Plotting Proportions

Sometimes we want proportions instead of counts.

```{r}
#| echo: true
ill_props <- oswego %>%
  count(ill) %>%
  mutate(
    prop = n / sum(n)
  )
```

Then plot:

```{r}
#| echo: true
ggplot(ill_props,aes(x = ill, y = prop)) +
  geom_col()
```

# Exercise: Modify the Code

Goal:

> Plot proportions of illness status rather than counts.

Start with:

```r
ill_counts <- oswego %>%
  count(ill)

ggplot(
  ill_counts,
  aes(x = ill, y = n)
) +
  geom_col()
```

Modify the code to create and plot a proportion.

# Exercise: Find the Error

What is wrong or confusing about this code?

```r
ill_counts <- oswego %>%
  count(ill)

ggplot(
  ill_counts,
  aes(x = ill)
) +
  geom_bar()
```

Hint:

`ill_counts` is already summarized.

# Checkpoint: Histograms and Bar Charts

## You should now be able to:

- Create histograms for numeric variables
- Adjust histogram bins
- Create bar charts for categorical or logical variables
- Explain what `geom_bar()` counts
- Use `geom_col()` with summarized data
- Plot counts or proportions

## Key ideas

- Histograms are for numeric distributions.
- Bar charts are for categorical counts.
- `geom_bar()` counts rows automatically.
- `geom_col()` uses precomputed y-values.

<div class="question">
Questions before we continue to scatterplots, boxplots, and density plots?
</div>


# Part III

## Scatterplots, Boxplots, and Density Plots

<div class="part">Comparing variables and comparing groups</div>

# Part III Learning Goals

By the end of this section, you should be able to:

- Create scatterplots
- Create boxplots
- Create density plots
- Choose an appropriate geom for a question
- Interpret common plot types
- Recognize when a variable is appropriate for a particular geom

# Choosing the Right Plot

A useful question is:

> What type of variable am I trying to visualize?

| Goal | Typical Plot |
|---|---|
| Distribution of one numeric variable | Histogram |
| Distribution of one numeric variable | Density plot |
| Relationship between two numeric variables | Scatterplot |
| Compare a numeric variable across groups | Boxplot |
| Compare counts across categories | Bar chart |

# Scatterplots

Scatterplots are used when:

- X is numeric
- Y is numeric

Goal:

- Explore relationships between variables
- Identify clusters
- Identify outliers
- Assess trends

# First Scatterplot

```{r}
#| echo: true
ggplot(oswego, aes(x = age, y = timesupper)) +
  geom_point()
```

Each point represents one attendee.

# Reading a Scatterplot

Questions to ask:

- Is there a positive relationship?
- Is there a negative relationship?
- Is there no obvious relationship?
- Are there unusual observations?

Scatterplots are often the first plot used when comparing two numeric variables.

# Exercise: Predict the Output

What do you think this will look like?

```r
ggplot(
  oswego,
  aes(
    x = age,
    y = timesupper
  )
) +
  geom_point()
```

Questions:

- What does each point represent?
- What variable is on each axis?

# Boxplots

Boxplots compare a numeric variable across groups.

Common examples:

- Age by illness status
- Blood pressure by treatment group

# First Boxplot

```{r}
#| echo: true
ggplot(oswego, aes(x = ill, y = age)) +
  geom_boxplot()
```

# What Does a Boxplot Show?

A boxplot summarizes:

- Median
- Lower quartile
- Upper quartile
- Potential outliers

It is a compact summary of a distribution.

# Interpreting the Median

The line inside the box is the median.

Questions:

- Is one group's median larger?
- Are the groups similar?
- Is there substantial variability?

# Boxplots and Outliers

Points outside the whiskers are often plotted separately.

These are potential outliers.

Outliers are not necessarily errors.

They are observations worth investigating.

# Boxplot + Jitter

A useful combination:

```{r}
#| echo: true
ggplot(oswego,aes(x = ill, y = age)) +
  geom_boxplot() +
  geom_jitter(
    width = 0.1, # how spread out points are on x-axis
    alpha = 0.5  # transparency
  )
```

This shows both:

- summary information
- individual observations

# Exercise: Predict the Output

What will this compare?

```r
ggplot(
  oswego,
  aes(
    x = milk,
    y = age
  )
) +
  geom_boxplot()
```

# Exercise: Write the Code

Goal:

> Compare `timesupper` across illness groups.

# Density Plots

Density plots are another way to visualize a distribution.

Example:

```{r}
#| echo: true
ggplot(oswego, aes(age)) +
  geom_density()
```

# Histogram vs Density Plot

Histogram:

```r
geom_histogram()
```

Density:

```r
geom_density()
```

Both display a distribution.

Density plots produce a smooth estimate.

# Why Use Density Plots?

Density plots can:

- Emphasize overall shape
- Remove dependence on bin selection
- Make group comparisons easier

# Overlaying Groups

Example:

```{r}
#| echo: true
ggplot(oswego, aes(x = age, color = ill)) +
  geom_density()
```

This creates one density curve for each illness group.

# Filled Density Curves

```{r}
#| echo: true
ggplot(oswego, aes(x = age, fill = ill)) +
  geom_density(alpha = 0.4)
```

Notice the use of:

```r
fill = ill
```

inside `aes()`.

# Common Mistake

Not all variables belong in every geom.

Example:

```{r}
#| echo: true
ggplot(oswego, aes(x = age, y = ill)) +
  geom_point()
```

This technically works.

But is it the best visualization?

Often a boxplot or density plot is more informative.

# Checkpoint: Scatterplots

You should now be able to:

- Create scatterplots
- Interpret points
- Use transparency
- Use jitter

## Key ideas

- Scatterplots compare two numeric variables.
- Each point is an observation.
- Transparency and jitter reduce overplotting.

<div class="question">
Questions before we move on?
</div>

# Checkpoint: Boxplots and Density Plots

You should now be able to:

- Create boxplots
- Create density plots
- Compare groups visually
- Choose an appropriate plot type

## Key ideas

- Boxplots summarize distributions.
- Density plots provide a smooth distribution estimate.
- Different plots answer different questions.
- Choosing the right plot is often more important than customizing it.

<div class="question">
Questions before we move to aesthetics?
</div>


# Part IV

## Aesthetics

<div class="part">Controlling how variables appear in a plot</div>

# Part IV Learning Goals

By the end of this section, you should be able to:

- Understand the purpose of `aes()`
- Map variables to visual properties
- Distinguish mapping from setting
- Use `fill`
- Use `color`
- Use `group`
- Interpret legends
- Diagnose common aesthetic mistakes

# What Are Aesthetics?

Aesthetics connect variables to visual features.

Examples:

```r
aes(
  x = age,
  fill = ill
)
```

Possible aesthetics include:

- `x`
- `y`
- `fill`
- `color`
- `group`
- `size`
- `shape`

# A Simple Mapping

```{r}
#| echo: true
ggplot(oswego,aes(x = age)) +
  geom_histogram()
```

The variable:

```r
age
```

is mapped to the x-axis.

# Mapping Additional Variables

```r
ggplot(
  oswego,
  aes(
    x = age,
    fill = ill
  )
) +
  geom_histogram()
```

Now illness status influences the appearance of the plot.

# A Key Idea

Variables belong inside:

```r
aes()
```

because they come from the dataset.

Example:

```r
aes(fill = ill)
```

Here:

```r
ill
```

is a variable.

# Mapping vs Setting

This is one of the most important concepts in ggplot.

Mapped:

```r
aes(fill = ill)
```

Set:

```r
fill = "steelblue"
```

These are not the same thing. Notice that aes is not used in the first example.

# Mapping Example

```{r}
#| echo: true
ggplot(oswego, aes(x = age, fill = ill)) +
  geom_histogram()
```

Result:

- Multiple groups
- Automatic legend
- Colors determined by illness status

# Setting Example

```{r}
#| echo: true
ggplot(oswego, aes(x = age)) +
  geom_histogram(fill = "steelblue")
```

Result:

- One color
- No grouping
- No legend

# Comparing the Two

Mapped:

```r
aes(fill = ill)
```

Question answered:

> Do the groups differ?

Set:

```r
fill = "steelblue"
```

Question answered:

> Show everything using the same color.

# Common Mistake

```r
aes(fill = "ill")
```

Notice the quotation marks.

This is usually wrong.

# Why Is It Wrong?

```r
aes(fill = ill)
```

means:

> Use the variable named ill.

But:

```r
aes(fill = "ill")
```

means:

> Treat the text "ill" as a category.

Very different behavior.

# Exercise: Find the Bug

What is wrong?

```r
ggplot(
  oswego,
  aes(
    x = age,
    fill = "vanilla"
  )
) +
  geom_histogram()
```

How would you fix it?

# Color vs Fill

Another common source of confusion.

```r
fill
```

controls interior color.

```r
color
```

controls outlines or lines.

# Fill Example

```{r}
#| echo: true
ggplot(oswego,aes(x = age,fill = ill)) +
  geom_histogram()
```

# Color Example

```{r}
#| echo: true
ggplot(oswego,aes(x = age,color = ill)) +
  geom_density()
```

The density curves receive different colors.

# When Do We Usually Use Fill?

Common examples:

```r
geom_bar()
geom_histogram()
geom_density()
geom_boxplot()
```

# When Do We Usually Use Color?

Common examples:

```r
geom_line()
geom_density()
geom_point()
```

# Legends

Mapped aesthetics usually create legends automatically.

Example:

```r
aes(fill = ill)
```

produces a legend describing the illness groups.

# No Mapping, No Legend

```r
geom_histogram(
  fill = "steelblue"
)
```

does not require a legend.

There is only one color.

# Using Multiple Aesthetics

```{r}
#| echo: true
ggplot(oswego,
  aes(
    x = age,
    fill = ill,
    color = ill
  )
) +
  geom_density(
    alpha = 0.3
  )
```

One variable can influence multiple visual properties.

# Grouping

Sometimes we want separate calculations for groups.

The `group` aesthetic controls this.

```r
aes(
  group = ill
)
```

# Why Group Matters

Imagine:

```r
ggplot(
  oswego,
  aes(
    x = age,
    color = ill
  )
) +
  geom_density()
```

Separate density curves are computed.

Grouping happens automatically.

# Explicit Grouping

Sometimes we specify it directly.

```{r}
#| echo: true
ggplot(
  oswego,
  aes(
    x = age,
    group = ill
  )
) +
  geom_density()
```

# Exercise: Mapping or Setting?

Should this go inside or outside `aes()`?

Question 1:

```r
ill
```

Question 2:

```r
"steelblue"
```

Question 3:

```r
sex
```

# Practical Rule

Ask yourself:

> Is this coming from a variable in my dataset?

If yes:

```r
aes(...)
```

If no:

Place it outside `aes()`.

# Example

Variable:

```r
aes(fill = ill)
```

Constant:

```r
fill = "steelblue"
```

# Common Error

```r
ggplot(
  oswego,
  aes(
    x = age
  )
) +
  geom_histogram(
    fill = ill
  )
```

Why does this fail?

Because:

```r
ill
```

is a variable and should be mapped.

# Another Common Error

```r
ggplot(
  oswego,
  aes(
    x = age,
    fill = ill
  )
) +
  geom_histogram(
    fill = "steelblue"
  )
```

Now the mapped value and set value compete.

The result is often confusing.

# Checkpoint: Aesthetics

You should now be able to:

- Explain what `aes()` does
- Distinguish mapping from setting
- Use `fill`
- Use `color`
- Use `group`
- Interpret legends

## Key Ideas

- Variables belong inside `aes()`.
- Constants belong outside `aes()`.
- Mapping creates groups and legends.
- Setting changes appearance without creating groups.

<div class="question">
Questions before we start creating grouped plots?
</div>

# Part V

## Grouped Plots and Position Adjustments

<div class="part">Comparing groups within a figure</div>

# Part V Learning Goals

By the end of this section, you should be able to:

- Create grouped bar charts
- Understand stacked bars
- Understand dodged bars
- Understand filled bars
- Choose an appropriate position adjustment

# Default Behavior

Going back to our bar chart from earlier:

```{r}
#| echo: true
ggplot(oswego,
  aes(
    x = sex,
    fill = ill
  )
) +
  geom_bar()
```

Bars are stacked by default.

# Position = "stack"

```r
geom_bar(
  position = "stack"
)
```

This is the default.

Each bar is divided into subgroups.

# Interpreting Stacked Bars

Questions answered:

- How many observations are there?
- What is the subgroup composition?

Questions that are harder:

- How do subgroup counts compare directly?

# Position = "dodge"

```{r}
#| echo: true
ggplot(
  oswego,
  aes(
    x = sex,
    fill = ill
  )
) +
  geom_bar(
    position = "dodge"
  )
```

Groups appear side-by-side.

# Why Use Dodge?

Dodged bars make it easier to compare counts across groups.

Example:

- Cases versus non-cases
- Within males and females

# Position = "fill"

```{r}
#| echo: true
ggplot(
  oswego,
  aes(
    x = sex,
    fill = ill
  )
) +
  geom_bar(
    position = "fill"
  )
```

Bars are scaled to the same height.

# What Does Fill Show?

Each bar represents:

```text
100%
```

The focus becomes:

- proportions
- composition

rather than counts. 

Notice the y-axis label should likely be changed when doing this.

# Comparing the Three

| Position | Best For |
|---|---|
| stack | Total counts and composition |
| dodge | Comparing counts |
| fill | Comparing proportions |

# Checkpoint: Grouped Plots

You should now be able to:

- Create grouped bar charts
- Use stack, dodge, and fill
- Interpret grouped displays

## Key Ideas

- Position adjustments change how groups are displayed.
- Different position choices answer different questions.
- There is rarely a single “correct” plot.

<div class="question">
Questions before we move to faceting?
</div>

# Part VI

## Faceting

<div class="part">Creating multiple plots from a single plotting command</div>

# Part VI Learning Goals

By the end of this section, you should be able to:

- Explain what faceting does
- Use `facet_wrap()`
- Use `facet_grid()`
- Choose appropriate faceting variables
- Interpret faceted plots
- Decide when faceting is preferable to color or fill

# What Is Faceting?

Faceting splits a dataset into groups and creates a separate panel for each group.

Instead of:

```r
one plot
```

you get:

```r
many related plots
```

with identical axes and (optionally) scales.

# Why Use Faceting?

Faceting is useful when:

- Comparing groups
- Looking for patterns within groups
- Avoiding cluttered plots
- Exploring interactions

# First Example

```{r}
#| echo: true
ggplot(oswego, aes(x = age)) +
  geom_histogram() +
  facet_wrap(~ sex)
```

# Reading Faceted Plots

Ask:

- Do the distributions differ?
- Are the centers similar?
- Is one group more variable?
- Are there outliers in only one group?

# Another Example

```{r}
#| echo: true
ggplot(oswego,
  aes(
    x = age,
    y = timesupper
  )
) +
  geom_point() +
  facet_wrap(~ ill)
```

Each panel contains a separate scatterplot.

# When Should You Facet?

Suppose you want to compare age distributions across illness groups.

Options:

```r
fill = ill
```

or

```r
facet_wrap(~ ill)
```

Both are reasonable.

Faceting often becomes easier when plots become crowded.

# Practical Advice

Good faceting variables:

- Sex
- Illness status
- Exposure status

Less useful:

- Variables with dozens of categories

Too many panels become difficult to read.

# facet_grid()

A second faceting approach:

```r
facet_grid(
  sex ~ ill
)
```

# How Does facet_grid Work?

Rows:

```r
sex
```

Columns:

```r
ill
```

Result:

A matrix of plots.

# Example

```{r}
#| echo: true
ggplot(oswego, aes(age)) +
  geom_histogram() +
  facet_grid(
    sex ~ ill
  )
```

# When Is facet_grid Useful?

Useful when you want to compare two grouping variables simultaneously.

Examples:

- Sex × Illness
- Treatment × Timepoint
- Region × Vaccination status

# Exercise

Modify:

```r
facet_wrap(~ sex)
```

to create panels using illness status.

# Exercise

Modify:

```r
facet_wrap(~ ill)
```

to create a grid using:

```r
sex
```

and

```r
ill
```

# Faceting vs Color

Color:

```r
aes(color = ill)
```

Advantages:

- Compact
- Single panel

Disadvantages:

- Can become cluttered

# Faceting vs Color

Faceting:

```r
facet_wrap(~ ill)
```

Advantages:

- Easier comparison
- Less overlap

Disadvantages:

- Requires more space

# Practical Rule

If curves overlap heavily:

Try:

```r
facet_wrap()
```

If there are only a few groups:

Try:

```r
color =
fill =
```

Both approaches are useful.

# Checkpoint: Faceting

You should now be able to:

- Use `facet_wrap()`
- Use `facet_grid()`
- Interpret faceted plots
- Choose between faceting and color grouping

## Key Ideas

- Faceting creates multiple panels.
- Panels represent subsets of the data.
- Faceting is often easier to interpret than overlapping groups.

<div class="question">
Questions before we start improving plots?
</div>

# Part VII

## Practical Plot Improvements

<div class="part">Making plots easier to read and easier to communicate</div>

# Part VII Learning Goals

By the end of this section, you should be able to:

- Add titles
- Add axis labels
- Choose a theme
- Flip coordinates
- Reorder categories
- Build a polished figure step-by-step

# Why Improve Plots?

Default plots are often useful.

But polished plots are:

- Easier to read
- Easier to interpret
- More professional
- Better for reports and presentations

# Plot Building Strategy

Start simple.

Then add:

1. Geometry
2. Labels
3. Theme
4. Additional refinements

# Step 1

```{r}
#| echo: true
ggplot(
  oswego,
  aes(age)
)
```

Not very useful yet.

# Step 2

```{r}
#| echo: true
ggplot(
  oswego,
  aes(age)
) +
  geom_histogram()
```

Now we have a figure.

# Step 3

```{r}
#| echo: true
ggplot(
  oswego,
  aes(age)
) +
  geom_histogram() +
  labs(
    title = "Age Distribution"
  )
```

# Step 4

```{r}
#| echo: true
ggplot(
  oswego,
  aes(age)
) +
  geom_histogram() +
  labs(
    title = "Age Distribution"
  ) +
  theme_classic()
```

# Adding Labels

Use:

```r
labs()
```

to add:

- Title
- Subtitle
- Axis labels
- Captions

# Example

```{r}
#| echo: true
ggplot(
  oswego,
  aes(age)
) +
  geom_histogram() +
  labs(
    title = "Age Distribution of Attendees",
    x = "Age (years)",
    y = "Count"
  )
```

# Why Labels Matter

Compare:

```r
x
y
```

versus

```r
Age (years)
Count
```

Clear labels reduce confusion.

# Themes

Themes control overall appearance.

Common examples:

```r
theme_classic()
theme_bw()
theme_minimal()
```

# theme_classic()

```{r}
#| echo: true
ggplot(
  oswego,
  aes(age)
) +
  geom_histogram() +
  theme_classic()
```

Produces a clean appearance.

# theme_bw()

```{r}
#| echo: true
ggplot(
  oswego,
  aes(age)
) +
  geom_histogram() +
  theme_bw()
```

Adds grid lines.

# Which Theme Should You Use?

There is no universally correct answer.

For most scientific figures:

```r
theme_classic()
```

is a reasonable default.

# Reordering Categories

Categorical variables often appear alphabetically.

Example:

```{r}
#| echo: true
food_long <- oswego %>%
  pivot_longer(
    cols = bakedham:fruitsalad,
    names_to = "food",
    values_to = "ate"
  )

head(food_long)
```

# Example Data

```{r}
#| echo: true
food_counts <- food_long %>%
  filter(ate) %>%
  count(food)
```

# Default Ordering

```{r}
#| echo: true
ggplot(food_counts, aes(x = food, y = n)) +
  geom_col()
```

Categories appear alphabetically.

# Reordering by Frequency

```{r}
#| echo: true
ggplot(food_counts,
  aes(
    x = reorder(food, n),
    y = n
  )) +
  geom_col()
```

Now foods are ordered by count.

# Why Reorder?

Questions become easier to answer.

Instead of:

```text
alphabetical order
```

we see:

```text
smallest → largest
```

or

```text
largest → smallest
```

# Flipping Coordinates

Long category labels often create crowded figures.

Use:

```r
coord_flip()
```

# Example

```{r}
#| echo: true
ggplot(food_counts, aes(x = reorder(food, n),y = n)) +
  geom_col() +
  coord_flip()
```

# Why Use coord_flip()?

Benefits:

- Easier to read labels
- Better for long category names
- Often preferred for ranked bar charts

# Exercise: Predict the Output

What changes when:

```r
coord_flip()
```

is added?

# Building a Complete Figure

Step 1:

```{r}
#| echo: true
ggplot(
  food_counts,
  aes(
    x = food,
    y = n
  )
)
```

# Step 2

```{r}
#| echo: true
ggplot(
  food_counts,
  aes(
    x = food,
    y = n
  )
) +
  geom_col()
```

# Step 3

```{r}
#| echo: true
ggplot(
  food_counts,
  aes(
    x = reorder(food, n),
    y = n
  )
) +
  geom_col()
```

# Step 4

```{r}
#| echo: true
ggplot(
  food_counts,
  aes(
    x = reorder(food, n),
    y = n
  )
) +
  geom_col() +
  coord_flip()
```

# Step 5

```{r}
#| echo: true
ggplot(
  food_counts,
  aes(
    x = reorder(food, n),
    y = n
  )
) +
  geom_col() +
  coord_flip() +
  labs(
    title = "Food Exposure Frequencies",
    x = "Food",
    y = "Number of Attendees"
  ) +
  theme_classic()
```

# Checkpoint: Improving Figures

You should now be able to:

- Add labels
- Add themes
- Reorder categories
- Flip coordinates
- Build plots incrementally

## Key Ideas

- Start simple.
- Improve plots step-by-step.
- Labels matter.
- Ordering categories often improves interpretation.
- Small improvements can dramatically improve readability.

<div class="question">
Questions before we start working with the food exposure data?
</div>


# Part VIII

## Food Exposure Visualizations

<div class="part">Using reshaped data to create richer plots</div>

# Part VIII Learning Goals

By the end of this section, you should be able to:

- Reshape food exposure variables into long format
- Count food exposures
- Create ranked bar charts
- Compare food exposure by illness status
- Plot counts and proportions
- Use food exposure data to build more informative visual summaries

# Why Reshape the Food Variables?

The Oswego dataset contains many food exposure variables.

Examples:

```r
bakedham
spinach
mashedpota
cabbagesal
jello
rolls
brownbread
milk
coffee
water
cakes
vanilla
chocolate
fruitsalad
```

Each variable indicates whether a person ate that food.

# Food Variables Are Wide

In the original dataset, each food is a separate column.

```text
id | ill | bakedham | spinach | milk | vanilla | chocolate
```

This format is useful for some analyses.

But for plotting many foods, it is often easier to use long format.

# Long Format for Food Exposures

Long format stores food names in one column and exposure status in another.

```text
id | ill | food      | ate
1  | TRUE | bakedham  | TRUE
1  | TRUE | spinach   | FALSE
1  | TRUE | vanilla   | TRUE
```

This structure is much easier to summarize and plot.

# Create `food_long`

In the previous section, we did this:

```r
food_long <- oswego %>%
  pivot_longer(
    cols = bakedham:fruitsalad,
    names_to = "food",
    values_to = "ate"
  )
```

This creates one row per person-food combination.

# What Does One Row Represent?

In `oswego`, one row is:

```text
one attendee
```

In `food_long`, one row is:

```text
one attendee-food combination
```

<div class="callout">
After pivoting, the number of rows increases because each person contributes multiple food rows.
</div>

# Exercise: Predict the Output

What will this count?

```r
food_long %>%
  count(food)
```

Questions:

- How many rows should there be?
- What does the `n` column represent?
- Does this tell us how many people ate each food?

# Counting Food Exposures

To count people who ate each food, first keep rows where `ate` is `TRUE`.

```r
food_counts <- food_long %>%
  filter(ate) %>%
  count(food)
```

This gives one row per food, with the number of attendees who ate it.

# Exposure Counts by Illness Status

Now we can count food exposures separately for cases and non-cases.

```{r}
#| echo: true
food_ill_counts <- food_long %>%
  filter(ate) %>%
  count(food, ill)
```

This creates counts for each combination of:

- food
- illness status

# Plot Exposure by Illness Status

```{r}
#| echo: true
ggplot(
  food_ill_counts,
  aes(
    x = reorder(food, n),
    y = n,
    fill = ill
  )
) +
  geom_col() +
  coord_flip()
```

This shows food exposure counts separated by illness status.

# Interpreting This Plot

Questions to ask:

- Which foods were commonly eaten?
- Among people who ate each food, how many became ill?
- Are some foods more common among cases?
- Are counts enough, or do we need proportions?

<div class="callout">
Counts can be useful, but outbreak investigations often require attack rates or proportions.
</div>

# Stacked vs Dodged Bars

Stacked bars:

```r
geom_col()
```

Dodged bars:

```r
geom_col(
  position = "dodge"
)
```

Dodged bars often make group comparisons easier.

# Dodged Food Exposure Plot

```{r}
#| echo: true
ggplot(
  food_ill_counts,
  aes(
    x = reorder(food, n),
    y = n,
    fill = ill
  )
) +
  geom_col(
    position = "dodge"
  ) +
  coord_flip()
```

# Proportions by Food

Counts are not always enough.

For each food, we may want the proportion of exposed attendees who became ill.

```{r}
#| echo: true
food_attack <- food_long %>%
  filter(ate) %>%
  group_by(food) %>%
  summarise(
    n_exposed = n(),
    attack_rate = mean(ill),
    .groups = "drop"
  )
```

Because `ill` is TRUE/FALSE, `mean(ill)` is the proportion ill.

# Plot Food-Specific Attack Rates

```{r}
#| echo: true
ggplot(
  food_attack,
  aes(
    x = reorder(food, attack_rate),
    y = attack_rate
  )
) +
  geom_col() +
  coord_flip()
```

This ranks foods by the proportion of exposed attendees who became ill.

# Add Labels

```{r}
#| echo: true
ggplot(
  food_attack,
  aes(
    x = reorder(food, attack_rate),
    y = attack_rate
  )
) +
  geom_col() +
  coord_flip() +
  labs(
    title = "Attack Rate by Food Exposure",
    x = "Food",
    y = "Proportion Ill Among Exposed"
  ) +
  theme_classic()
```

# Counts and Rates Answer Different Questions

Counts answer:

```text
How many exposed attendees became ill?
```

Rates answer:

```text
Among exposed attendees, what proportion became ill?
```

Both can be useful.

They are not the same.

# Faceting Food Plots

Faceting can help compare plots by illness status.

```{r}
#| echo: true
food_long %>%
  filter(ate) %>%
  count(food, ill) %>%
  ggplot(
    aes(
      x = reorder(food, n),
      y = n
    )
  ) +
  geom_col() +
  coord_flip() +
  facet_wrap(~ ill)
```

# What Does the Faceted Plot Show?

Each panel corresponds to one illness group.

Within each panel:

- foods are shown
- counts are displayed
- food labels are easier to compare than in a crowded stacked plot

# A More Complete Food Plot

```{r}
#| echo: true
food_attack %>% ggplot(aes(x = reorder(food, attack_rate), y = attack_rate)) +
  geom_col() +
  coord_flip() +
  labs(
    title = "Attack Rate by Food Exposure",
    subtitle = "Among attendees who ate each food",
    x = "Food",
    y = "Proportion Ill"
  ) +
  theme_classic()
```

# Interpreting Food Exposure Plots

When interpreting food plots, consider:

- How many people ate each food?
- What proportion became ill?
- Are there small denominators?
- Are missing values present?
- Is the association plausible?

Visualization is a starting point, not the final analysis.

# Checkpoint: Food Exposure Visualizations

You should now be able to:

- Reshape food columns into long format
- Count food exposures
- Plot ranked food exposure counts
- Plot food exposure by illness status
- Compute and plot food-specific attack rates

## Key Ideas

- Long format makes many plots easier.
- Counts and proportions answer different questions.
- `mean(ill)` works because `ill` is TRUE/FALSE.
- Reordering and flipping coordinates often improve food exposure plots.

<div class="question">
Questions before we move to common ggplot errors?
</div>


# Part IX

## Common ggplot Errors

<div class="part">Learning to read and fix common mistakes</div>

# Learning Goals

- Recognize common ggplot errors
- Interpret error messages
- Debug plots systematically
- Understand why mistakes occur

# Error: Missing Aesthetic

```r
ggplot(oswego) +
  geom_histogram()
```

Typical message:

```text
stat_bin() requires an x or y aesthetic
```

Why?

A histogram needs a numeric variable.

# Fix

```r
ggplot(
  oswego,
  aes(x = age)
) +
  geom_histogram()
```

# Error: Object Not Found

```r
ggplot(
  oswego,
  aes(x = agge)
) +
  geom_histogram()
```

Common causes:

- Typo
- Wrong variable name
- Variable not in data frame

# Diagnose

```r
names(oswego)
```

# Error: Forgot the Plus Sign

```r
ggplot(
  oswego,
  aes(age)
)
geom_histogram()
```

Each layer must be connected using `+`.

# Error: Quoted Variable

Wrong:

```r
aes(fill = "ill")
```

Correct:

```r
aes(fill = ill)
```

# Why?

Inside `aes()`:

```r
ill
```

means variable.

```r
"ill"
```

means literal text.

# Error: Wrong Geom

```r
ggplot(
  oswego,
  aes(age)
) +
  geom_bar()
```

This usually isn't what you want.

For numeric distributions:

```r
geom_histogram()
```

is often more appropriate.

# Error: Using Variables Outside aes()

Wrong:

```r
geom_histogram(
  fill = ill
)
```

Correct:

```r
aes(fill = ill)
```

# Debugging Strategy

1. Read the error
2. Check variable names
3. Check aesthetics
4. Run code one layer at a time
5. Build plots incrementally

# Exercise

Identify the bug:

```r
ggplot(
  oswego,
  aes(
    x = age,
    fill = "ill"
  )
) +
  geom_histogram()
```

# Checkpoint

You should now be able to:

- Read common error messages
- Diagnose mapping problems
- Fix common ggplot mistakes




# Part X

## Tables with gt

<div class="part">Creating polished tables</div>

# Learning Goals

- Create gt tables
- Format numbers
- Rename columns
- Add titles
- Improve readability

# Why gt?

`gt` creates publication-quality tables directly from data frames.

# Basic Table

```{r}
#| echo: true
library(gt)

ill_counts <- oswego %>%
  count(ill)

ill_counts %>%
  gt()
```

# Column Labels

```{r}
#| echo: true
ill_counts %>%
  gt() %>%
  cols_label(
    ill = "Illness Status",
    n = "Count"
  )
```

# Formatting Numbers

```{r}
#| echo: true
ill_counts %>%
  gt() %>%
  fmt_number(
    columns = n,
    decimals = 0
  )
```

# Table Titles

```{r}
#| echo: true
ill_counts %>%
  gt() %>%
  tab_header(
    title = "Illness Counts"
  )
```

# Combining Steps

```{r}
#| echo: true
ill_counts %>%
  gt() %>%
  cols_label(
    ill = "Illness Status",
    n = "Count"
  ) %>%
  tab_header(
    title = "Illness Counts"
  )
```

# Percentages

```{r}
#| echo: true
ill_props <- oswego %>%
  count(ill) %>%
  mutate(
    prop = n / sum(n)
  )
```

# Format Percentages

```{r}
#| echo: true
ill_props %>%
  gt() %>%
  fmt_percent(
    columns = prop,
    decimals = 1
  )
```

# Day 2 Cheat Sheet

## Core ggplot Functions

```r
ggplot()
aes()

geom_histogram()
geom_bar()
geom_col()
geom_point()
geom_boxplot()
geom_density()

facet_wrap()
facet_grid()

labs()

theme_classic()
theme_bw()

coord_flip()
```

# Common Wrangling Used for Plots

```r
count()

group_by()

summarise()

pivot_longer()

mutate()

filter()
```

# Table Functions

```r
gt()

cols_label()

fmt_number()

fmt_percent()

tab_header()
```

# gtsummary Functions

```r
tbl_summary()

add_p()

as_gt()
```

# Final Takeaways

You should now be able to:

- Build common statistical graphics
- Interpret and customize plots
- Debug common ggplot problems
- Create gt tables
- Communicate results visually and reproducibly

# Looking Ahead

The skills from Days 1 and 2 combine:

```r
Import Data
→ Wrangle Data
→ Visualize Data
→ Summarize Results
→ Communicate Findings
```

Questions?


