---
title: "Final Practice: Reproducible Survey Analysis"
subtitle: "R Bootcamp"
author: "Your Name"
format:
  html:
    toc: true
    toc-depth: 2
    number-sections: true
execute:
  echo: true
  warning: false
  message: false
---

# Background

You are working with data from a synthetic public health survey. The survey was administered to residents in several counties after a possible community exposure event.

The data are stored in three separate files:

- `survey_respondents.csv`: one row per person in the respondent roster
- `survey_health.csv`: health symptoms, care-seeking, and testing information
- `survey_exposures.csv`: possible exposure information

These files are intentionally separated so that you can practice joining related datasets.

Some survey questions use skip logic. For example, respondents who did not report symptoms were not asked symptom follow-up questions. Respondents who did not eat at a restaurant were not asked restaurant follow-up questions.

For those follow-up questions, missing values may occur for two different reasons:

- `"not_asked"`: the question was skipped by design
- `"non_response"`: the question was asked, but the respondent did not answer
- `NA`: the question was answered

Your goal is to create a short reproducible Quarto report that imports the data, checks the joins, summarizes key variables, investigates missingness, and communicates a few findings with tables and figures.

# Setup

```{r}
library(tidyverse)
library(here)
library(gt)
library(gtsummary)
```

# 1. Import the data

Read the three survey files from the `Data` folder.

```{r}
respondents <- read_csv(
  here("Data", "survey_respondents.csv")
)

health <- read_csv(
  here("Data", "survey_health.csv")
)

exposures <- read_csv(
  here("Data", "survey_exposures.csv")
)
```

## Question 1.1

How many rows and columns are in each dataset?

```{r}
# Use dim(), nrow(), ncol(), or glimpse() here

```

**Write your interpretation here:**

> 

## Question 1.2

What does one row represent in each dataset?

**Write your answer here:**

> 

# 2. Inspect the data

Before joining datasets, inspect the variable names and data structures.

```{r}
# Inspect each dataset

```

## Question 2.1

What variable appears to link the three datasets together?

**Write your answer here:**

> 

## Question 2.2

Which variables are logical, categorical, numeric, or dates?

**Write your notes here:**

> 

# 3. Check overlap across files

Not everyone completed every survey component. Use joins to identify respondents who appear in one file but not another.

## Question 3.1

Which respondents are in the roster but missing from the health survey?

```{r}
# Use anti_join()

```

**Write your interpretation here:**

> 

## Question 3.2

Which respondents are in the roster but missing from the exposure survey?

```{r}
# Use anti_join()

```

**Write your interpretation here:**

> 

## Question 3.3

How many respondents completed all three survey components?

```{r}
# One approach:
# respondents %>%
#   inner_join(health, by = "respondent_id") %>%
#   inner_join(exposures, by = "respondent_id") %>%
#   nrow()

```

**Write your interpretation here:**

> 

# 4. Create an analysis dataset

Create one dataset that starts with the respondent roster and attaches health and exposure information.

```{r}
survey <- respondents %>%
  left_join(health, by = "respondent_id") %>%
  left_join(exposures, by = "respondent_id")
```

## Question 4.1

How many rows are in the joined dataset? Is this what you expected?

```{r}
# Check number of rows

```

**Write your interpretation here:**

> 

## Question 4.2

Why might `left_join()` be a reasonable choice here?

**Write your answer here:**

> 

# 5. Basic respondent summaries

Summarize the respondent roster.

## Question 5.1

How many respondents are in each county?

```{r}
# Count respondents by county

```

## Question 5.2

What is the age distribution of respondents?

```{r}
# Summarize age using mean, median, range, etc.

```

## Question 5.3

Create a figure showing the age distribution.

```{r}
# Create a histogram of age

```

**Write 1–2 sentences interpreting the figure:**

> 

# 6. Health outcomes

## Question 6.1

What proportion of respondents reported symptoms?

```{r}
# Since had_symptoms is TRUE/FALSE, mean(had_symptoms, na.rm = TRUE) gives a proportion

```

**Write your interpretation here:**

> 

## Question 6.2

Among respondents who reported symptoms, which symptoms were most common?

```{r}
# One possible approach:
# survey %>%
#   filter(had_symptoms) %>%
#   summarise(
#     fever = mean(fever, na.rm = TRUE),
#     cough = mean(cough, na.rm = TRUE),
#     nausea = mean(nausea, na.rm = TRUE),
#     diarrhea = mean(diarrhea, na.rm = TRUE)
#   )

```

**Write your interpretation here:**

> 

## Question 6.3

Create a plot of symptom frequencies among symptomatic respondents.

```{r}
# Hint:
# 1. Filter to had_symptoms == TRUE
# 2. Pivot fever, cough, nausea, diarrhea into long format
# 3. Keep rows where symptom value is TRUE
# 4. Count symptom
# 5. Plot with geom_col()

```

**Write 1–2 sentences interpreting the figure:**

> 

# 7. Missingness and skip logic

Some follow-up questions are missing because they were not asked. Others are missing because of nonresponse.

## Question 7.1

For the fever question, how many missing values are due to `not_asked` versus `non_response`?

```{r}
# Count fever_missing_reason

```

**Write your interpretation here:**

> 

## Question 7.2

Repeat this for restaurant name missingness.

```{r}
# Count restaurant_name_missing_reason

```

**Write your interpretation here:**

> 

## Question 7.3

Why is it important to distinguish `not_asked` from `non_response`?

**Write your answer here:**

> 

# 8. Exposure summaries

## Question 8.1

What proportion of respondents attended the event, ate at a restaurant, drank tap water, or had contact with a sick person?

```{r}
# Summarize exposure variables using mean(..., na.rm = TRUE)

```

**Write your interpretation here:**

> 

## Question 8.2

Create a plot showing exposure frequencies.

```{r}
# Hint:
# Pivot attended_event, ate_restaurant, drank_tap_water,
# and contact_with_sick_person into long format.
# Then count exposures where value is TRUE.

```

**Write 1–2 sentences interpreting the figure:**

> 

# 9. Health outcomes by exposure

Choose one exposure and compare symptom reporting between exposed and unexposed respondents.

## Question 9.1

Compare the proportion with symptoms by restaurant exposure.

```{r}
# Group by ate_restaurant and summarize mean(had_symptoms, na.rm = TRUE)

```

**Write your interpretation here:**

> 

## Question 9.2

Create a figure showing symptom proportion by restaurant exposure.

```{r}
# Create a summarized dataset first, then plot with geom_col()

```

**Write 1–2 sentences interpreting the figure:**

> 

# 10. Summary table

Create a descriptive table comparing respondents with and without symptoms.

```{r}
# Try this:
# survey %>%
#   select(age, sex, county, attended_event, ate_restaurant,
#          drank_tap_water, contact_with_sick_person, had_symptoms) %>%
#   tbl_summary(by = had_symptoms)

```


# 12. Optional challenge

Create one polished figure and one polished table that you would include in a short report.

## Figure

```{r}
# Your polished figure here

```

## Table

```{r}
# Your polished table here

```

# Wrap-up

In this exercise, you practiced:

- Importing data
- Inspecting data
- Joining files
- Checking unmatched records
- Working with skip-logic missingness
- Summarizing logical variables
- Pivoting survey questions
- Creating plots
- Creating tables
- Writing a reproducible Quarto report
